CS 2051 Group Project

### 2025-04-06 latex/paper.tex
Added names
Small change adding names to the author tag in the paper

\documentclass{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{geometry}
\geometry{legalpaper, portrait, margin=.5in}
\title{CS 2051 Group Project}
\author{}
\date{March 2025}

@@

\maketitle
\end{document}

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{graphicx}

\geometry{legalpaper, portrait, margin=.5in}

\title{CS 2051 Group Project}
\author{Ibrahim Ahmad, Cipriano Dorbessan, Christopher Gossmann}
\date{April 2025}

@@

\maketitle

\section*{Project Outline}
Our project is focused on using Newton’s method (also known as the Newton-Raphson method) as the optimizer for machine learning. We will use \LaTeX{} to make our research paper. Since our project involves a significant amount of coding, we created a GitHub where we will keep our code: \url{https://github.com/cipridorbe/Newtons-Method}

\section*{Background Knowledge}

\subsection*{Newton’s Method}
Newton’s method is an iterative root-finding algorithm for functions. For a single variable function \( f: \mathbb{R} \to \mathbb{R} \), the algorithm starts with an initial guess \( x_0 \), and the next term \( x_1 \) is found by using the tangent line at \( x_0 \).

The tangent through \( (x_0, f(x_0)) \) with slope \( f'(x_0) \) is given by:
\[
y - f(x_0) = f'(x_0)(x - x_0)
\]
Solving \( y = 0 \), we get:
\[
x = x_0 - \frac{f(x_0)}{f'(x_0)}
\]
So the formula for the \(n\)-th term is:
\[
x_{n+1} = x_n - \frac{f(x_n)}{f'(x_n)}
\]

In machine learning, we aim to find the minimum of a function, not its root. Since the minimum occurs when \( f'(x) = 0 \), we can use Newton’s method to find the minimum:
\[
x_{n+1} = x_n - \frac{f'(x_n)}{f''(x_n)}
\]

For multivariate functions \( f: \mathbb{R}^k \to \mathbb{R} \), we generalize the derivative to the gradient and the second derivative to the Hessian matrix. Thus, the update formula becomes:
\[
x_{n+1} = x_n - [\nabla^2 f(x_n)]^{-1} \nabla f(x_n)
\]
Where \( x_n, x_{n+1}, \nabla f(x_n) \in \mathbb{R}^k \), and \( \nabla^2 f(x_n) \in \mathbb{R}^{k \times k} \).

Computing the inverse of the Hessian matrix is expensive. Instead, we solve the linear system:
\[
[\nabla^2 f(x_n)] h = -\nabla f(x_n)
\]
Which gives \( h = -[\nabla^2 f(x_n)]^{-1} \nabla f(x_n) \).

\section*{Differentiation}
To compute derivatives, we can use:

\begin{itemize}
    \item \textbf{Finite differencing:} Simple to implement but computationally expensive when \( f''(x_n) \) is needed many times.
    \item \textbf{Automatic differentiation:} Builds a computational graph to calculate derivatives. It is complex to implement but more efficient in higher-order systems.
\end{itemize}

\section*{Method and Complexity in Machine Learning}
To apply Newton’s method in machine learning, we require:
\begin{itemize}
    \item An initial guess (often random).
    \item A tolerance for convergence.
    \item A maximum number of iterations.
\end{itemize}

Newton’s method converges quadratically near the solution — each iteration roughly doubles the number of correct digits. However, poor initial guesses may lead to slow convergence or divergence. The function must have continuous first and second derivatives.

Performance also depends on the function shape. Linear first derivatives tend to lead to faster convergence.

Although Newton’s method has a quadratic convergence rate, it can have greater complexity than other optimization algorithms due to:

\begin{itemize}
    \item Gradient computation: \( \mathcal{O}(n) \)
    \item Hessian computation: \( \mathcal{O}(n^2) \)
    \item Solving linear systems: \( \mathcal{O}(n^3) \) (for dense matrices)
\end{itemize}

The number of iterations varies depending on the tolerance, curvature, and initial guess.

\section*{Conclusion}
We have not yet implemented and benchmarked these algorithms, so the efficiency remains unknown.

\section*{Extensions and Applications}
Potential extensions:
\begin{itemize}
    \item Apply Newton’s method to other domains: physics engines, architectural design, etc.
    \item Use in other machine learning fields: unsupervised or reinforcement learning.
    \item Explore other optimization algorithms, including derivative-free methods.
\end{itemize}

If results are promising, we aim to test on larger networks and frameworks.

\section*{References}
\begin{enumerate}
    \item Wikipedia, “Newton’s method.” \url{https://en.wikipedia.org/wiki/Newton%27s_method}
    \item GeeksforGeeks, “Newton’s method in machine learning.” \url{https://www.geeksforgeeks.org/newtons-method-in-machine-learning/}
    \item Wikipedia, “Automatic differentiation.” \url{https://en.wikipedia.org/wiki/Automatic_differentiation}
\end{enumerate}

\end{document}
